From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/discussion.tex

%%%%%%%%%%%%%%
\paragraph{Rewriting} Term-rewriting systems and word problems have a long history of being formalised in proof
assistants. In the recent past, higher order rewriting systems have been formalised in {homotopy.io}, Lean, Coq, and
HoTT~\citet{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main result, we developed a rewriting
system for the Coxeter relations for $\Sn$ to solve its word problem and formalised its correctness and termination in a
proof assistant. Our design based on encoding permutations as adjacent transpositions led to an intuitive whose
correctness was straightforward to establish and whose termination only required slightly modified variations on
Church-Rosser theorems. Other encodings of permutations as listed vectors, matrices, inductively generated trees
(Motzkin trees), Young diagrams, or string diagrams, proved either difficult formalise in type theory or difficult to
relate directly to the primitive type isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970}
produced too many equations making proving correctness and termination intractable.

>>>>

%%%%%%%%%%%%%%
\paragraph{Rewriting} Higher-order term-rewriting systems and word problems have a long history of being formalised in
proof assistants like {homotopy.io}, Lean, Coq, and HoTT~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the
proof of our main result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem
and formalised its correctness and termination in Agda. Our design based on encoding permutations as
adjacent transpositions led to an intuitive system whose correctness was straightforward to establish and whose termination
only required slightly modified variations on Church-Rosser theorems. Other encodings of permutations as listed vectors,
matrices, inductively generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult
formalise in type theory or difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
\citet{knuthSimpleWordProblems1970} completion produced too many equations making proving correctness and termination
intractable.

### 2021-07-09 Pi+/paper/ufin.tex

  Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, squashes the elements of $A$ turning it into a
  proposition. It is given by a HIT with a point constructor $\trunc{\blank} : A \to \Trunc{A}$, and a path constructor
  $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$, which equates every pair of points in the truncation
  (see~\Cref{def:prop-trunc}).

@@

  Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes an $\hSet$ $A$ and a relation $R : A \to A \to
    \UU$. It has an inclusion of points $q : A \to \quot{A}{R}$, and adds paths between related pairs of elements
  $\quotrel : R(x,y) \to q(x) \id_{\quot{A}{R}} q(y)$ (see~\Cref{def:set-quot}).  We recall that the quotient is
  \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x) \id_{\quot{A}{R}}
    q(y))$.

@@

  \begin{definition}
    \label{app:defone}

@@

    \label{app:propone}

@@

>>>>

  Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, squashes the elements of $A$ turning it into a
  proposition. It is given by a HIT with a point constructor $\trunc{\blank} : A \to \Trunc{A}$, and a path constructor
  $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$, which equates every pair of points in the truncation
  (see~\cref{def:prop-trunc}).

@@

  \begin{definition}[${\fib_{f}} : B \to \UU$]
    \label{def:fib}
    The fiber of $f : A \to B$ at $b : B$ is
    \[
      \fib_{f}(b) \defeq \dsum{a:A}{f(a) \id_{B} b}.
    \]
  \end{definition}

  \begin{definition}[${\term{im}} : (f : A \to B) \to \UU$]
    \label{def:im}
    The image of $f$ is the (-1)-truncation of its fiber.
    \[
      \im{f} \defeq \dsum{b:B}{\Trunc[-1]{\fib_{f}(b)}}
    \]
  \end{definition}

  \begin{proposition}
    The following are equivalent.
    \begin{enumerate}
      \item $f : A \to B$ is an equivalence.
      \item $f$ has a left and right inverse.
      \item $f$ has contractible fibers.
    \end{enumerate}
  \end{proposition}

  Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes an $\hSet$ $A$ and a relation $R : A \to A \to
    \UU$. It has an inclusion of points $q : A \to \quot{A}{R}$, and adds paths between related pairs of elements
  $\quotrel : R(x,y) \to q(x) \id_{\quot{A}{R}} q(y)$ (see~\cref{def:set-quot}).  We recall that the quotient is
  \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x) \id_{\quot{A}{R}}
    q(y))$.

@@

  \begin{definition}[Additive symmetric monoidal structure]
    \label{def:additive}

@@

    \label{prop:additive}

@@

  \begin{definition}[Multiplicative symmetric monoidal structure]
    \label{def:multiplicative}
      I                 & \defeq F_{1}                                           \\
      X \otimes Y       & \defeq X \times Y                                      \\
      \lambda_{X}       & : I \times X \eqv X                                    \\
      \rho_{X}          & : X \times I \eqv X                                    \\
      \alpha_{X,Y,Z}    & : (X \otimes Y) \otimes Z \eqv X \otimes (Y \otimes Z) \\
      \mathcal{B}_{X,Y} & : X \otimes Y \eqv Y \otimes X
    \label{prop:multiplicative}
        {(X \otimes I) \otimes Y} && {X \otimes (I \otimes Y)} \\
        {} & {X \otimes Y}
        \arrow[""{name=0, anchor=center, inner sep=0}, "{\rho_{X} \otimes 1_{Y}}"', from=1-1, to=2-2]
        \arrow[""{name=1, anchor=center, inner sep=0}, "{1_{X} \otimes \lambda_{Y}}", from=1-3, to=2-2]
        & {(W \otimes X) \otimes (Y \otimes Z)} \\
        {((W \otimes X) \otimes Y) \otimes Z} && {W \otimes (X \otimes (Y \otimes Z))} \\
        {(W \otimes (X \otimes Y)) \otimes Z} && {W \otimes ((X \otimes Y) \otimes Z)}
        \arrow["{\alpha_{W \otimes X, Y, Z}}", from=2-1, to=1-2]
        \arrow["{\alpha_{W,X,Y \otimes Z}}", from=1-2, to=2-3]
        \arrow["{1_{W} \otimes \alpha_{X,Y,Z}}"', from=4-3, to=2-3]
        \arrow["{\alpha_{W,X,Y} \otimes 1_{Z}}"', from=2-1, to=4-1]
        \arrow["{\alpha_{W,X \otimes Y,Z}}"', from=4-1, to=4-3]
        & {X \otimes (Y \otimes Z)} \\
        {(X \otimes Y) \otimes Z} && {(Y \otimes Z) \otimes X} \\
        {(Y \otimes X) \otimes Z} && {Y \otimes (Z \otimes X)} \\
        & {Y \otimes (X \otimes Z)}
        \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y} \otimes 1_{Z}}"', from=2-1, to=3-1]
        \arrow["{1_{Y} \otimes \mathcal{B}_{X,Z}}"', from=4-2, to=3-3]
        \arrow["{\mathcal{B}_{X,Y \otimes Z}}", from=1-2, to=2-3]
        {X \otimes Y} && {X \otimes Y} \\
        & {Y \otimes X}
        \arrow["{1_{X \otimes Y}}", Rightarrow, no head, from=1-1, to=1-3]

\begin{toappendix}
  \begin{proposition}[Distributivity]
    \label{prop:distributivity}
    \begin{gather*}
      \begin{aligned}
        \delta_{l} : X \otimes (Y \oplus Z) & \eqv (X \otimes Y) \oplus (X \otimes Z) \\
        \delta_{r} : (X \oplus Y) \otimes Z & \eqv (X \otimes Z) \oplus (Y \otimes Z)
      \end{aligned}
      \qquad
      \begin{aligned}
        a_{l} : X \otimes O \eqv O \\
        a_{r} : O \otimes X \eqv O
      \end{aligned}
    \end{gather*}
  \end{proposition}
\end{toappendix}
